---
title: "Data Exercise 1"
author: "Bob Castelein"
date: "17-5-2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Netherlands' banana trade


## Introduction

One of the main cargoes shipped in reefer containers is bananas. They make up some 20% of all conditioned cargo, and impose some of the most stringent requirements on energy security, due to the narrow temperature bandwidth (2 degrees at most) at which the product should be preserved.


## Collecting the data

In this exercise, I will work with data of the Netherlands' import and export flows of bananas. From the FAO statistics website (http://www.fao.org/faostat/en/#home) I pull the trade matrix data for the Netherlands as the reporting country, selecting all partner countries, and bananas as the item, for the years 2000-2013. This raw data file is available for download at my github page.

First I download the data and store it in a variable called 'bananas'.

```{r}
bananas <- read.csv(file = "https://raw.githubusercontent.com/bobcastelein/session2/master/nlbananas2000-2013.csv")
head(bananas)
```


## Data cleaning

Right now, the data looks a bit messy, with several variables we don't need. We are interested in the partner country (the reporting country is always the Netherlands), the element (export quantity/value, and import quantity/value - although this distinction is not as useful when considering one commodity only), the year, and the value. To work with these variables alone, I use the dplyr package to select the appropriate variables.

```{r, message = FALSE}
library(dplyr)
invisible(bananasl <- bananas %>% select(Partner.Countries, Element, Year, Unit, Value))
head(bananasl)
```

The data is still in a long format (various rows per country per element, denoted by bananasl), so I transform it to a wide format. First I unite the element and unit columns into one variable called 'flow'.

```{r}
library(tidyr)
invisible(bananasl <- bananasl %>% unite(flow, Element, Unit, sep = "_"))
head(bananasl)
```

To transform it to a wide format, I 'spread' the flow and value variables, and rename the new variables to facilitate coding and reading.

```{r}
invisible(bananasw <- bananasl %>% spread(flow, Value))
invisible(bananasw <- bananasw %>% rename(exquant = `Export Quantity_tonnes`, exval = `Export Value_1000 US$`, imquant = `Import Quantity_tonnes`, imval = `Import Value_1000 US$`))
head(bananasw)
```


Subsequently I rank the countries according to mean import and export value over the period 2000-2013, and select the 10 most important import and export partners.

```{r}
head(bananasw %>% group_by(Partner.Countries) %>% summarize(mean(imval)) %>% arrange(desc(`mean(imval)`)), 10)
head(bananasw %>% group_by(Partner.Countries) %>% summarize(mean(exval)) %>% arrange(desc(`mean(exval)`)), 10)
```

For the banana trade between the Netherlands and these countries I want to plot the developments in import and export volumes, expressed in total import/export value. First I select only those rows that correspond with the countries in these top 10s.

```{r}
topexport <- bananasw %>% group_by(Partner.Countries) %>% mutate(mean(exval)) %>% filter(mean(exval) > 2182)
topimport <- bananasw %>% group_by(Partner.Countries) %>% mutate(mean(imval)) %>% filter(mean(imval) > 2504)
topimport
topexport
```

Now we have two data frames, one with the export data for the Netherlands' 10 most important banana export partners, and one with the import data for the Netherlands' 10 most important banana import partners.


## Visualization and discussion

Below we can create two plots: one of the volume imported from the 10 most important import partners over the period 2000-2013, and one of the volume exported to the 10 most important export partners over the period 2000-2013. 

```{r}
library(ggplot2)
ggplot(topimport, aes(x = Year, y = imval, colour = factor(Partner.Countries))) + geom_point() + geom_line() + labs(x = "Year", y = "Value of banana imports ($1000)", colour = "Import origins", title = "Banana imports from 10 major import partners, 2000-2013") + scale_y_continuous(breaks = c(0, 25000, 50000, 75000, 100000), limits = c(0, 100000))
```

The first thing we notice is that for a long time, Belgium was the Netherlands' primary source of bananas. As bananas don't grow in Belgium, the obvious conclusion is that bananas were imported in Belgium re-exported to the Netherlands. Over time this volume has shrunk, and the the value of banana imports from major banana producers such as Colombia, Ecuador, Peru, and Costa Rica have increased. The world market for fruits is dominated by a few very large players (Chiquita, Bonita, Fyffes, Dole etc.), who have a large extent of control over their global supply chains, from farmer to supermarket. The increase in flows from banana-producing countries, and a decrease in the flow from Belgium (traditionally a main player on transatlantic routes through the Port of Antwerp), is an indication that the Dutch transportation sector has been successful in attracting direct cargo flows from the major banana traders. The rise of Germany as an important source of banana imports reflects the competition between Northern European ports for largely the same hinterlands. 

```{r, message = FALSE, warning = FALSE}
ggplot(topexport, aes(x = Year, y = exval, colour = factor(Partner.Countries))) + geom_point() + geom_line() + labs(x = "Year", y = "Value of banana exports ($1000)", colour = "Export destinations", title = "Banana exports to 10 major export partners, 2000-2013") + scale_y_continuous(breaks = c(0, 25000, 50000, 75000, 100000), limits = c(0, 100000))
```

As the climate in the Netherlands does not suit itself to grow bananas, the bananas imported are either consumed domestically or re-exported. The main export market nowadays (at least as of 2013) is Germany, with other countries such as the United Kingdom, Poland, France and Belgium making up smaller markets. 